---
title: Correlation, linear regression, and Statistical Significance between Conductivity
  and Total Dissolved Solids
Name: Christian Shaffer
output: pdf_document
  pdf_document: default
  html_notebook: default
---

```{r}
# Import the CSV
water_stats <- read.csv("water_potability.csv")

# Summary of CSV imported
summary(water_stats)

# First 5 rows of CSV
head(water_stats, 5)

# Check data type
str(water_stats)
```

## Create sample size of population.

```{r}
# For reproducibility
set.seed(1)

# Randomly select 50 numbers to represent rows in dataframe
sampleIDs <- sample(1:3276, size = 50, replace = FALSE)

# Select the columns I'm interested in
myColumns <- c("Conductivity", "Solids")

# Store the randomly selected rows on the given columns
sample_water <- water_stats[sampleIDs, myColumns]
sample_water
```

## Descriptive Statistics of Conductivity within Dataset.

```{r}
# Statistical Model for calculating base statistical stats from Data Frame, specifically Turbidity and Solids

# Calculate the mean, median, variance, and standard deviation:

mu_con <- mean(water_stats$Conductivity, na.rm = T) # calculates the population mean, using mu

median_con <- median(water_stats$Conductivity, na.rm = T) # calculates the median, using na.rm to remove NA's

diff_con <- water_stats$Conductivity - mu_con # deviance between data frame and mean (X - xbar)

Sqr <- diff_con^2 # squared of the deviance (X - xbar)^2

SOS <- sum(Sqr) # Sum of squares using square of the deviance

var_con <- var(water_stats$Conductivity) # calculates the variance

myvar_con <- SOS / length(water_stats$Conductivity) # calculates the variance from difference between SOS and number of scores(data points)

sd_con <- sd(water_stats$Conductivity) # calculates sample the standard deviation
 
mu_con <- sqrt(myvar_con) # calculates the population standard deviation given by sigma

# The range calculated two ways:

range_con <- diff(range(water_stats$Conductivity)) # calculates the range with the diff function
  
min_con <- min(water_stats$Conductivity) # minimum value

max_con <- max(water_stats$Conductivity) # maximum value

# range_turb <- max_turb - min_turb # range of data frame using min and max functions

```

## Descriptive Statistics of Solids within Dataset.

```{r}
# Statistical Model for calculating base statistical stats from Data Frame, specifically Conductivity and Solids

# Calculate the mean, median, variance, and standard deviation:

mu_solids <- mean(water_stats$Solids, na.rm = T) # calculates the population mean, using mu

median_solids <- median(water_stats$Solids, na.rm = T) # calculates the median, using na.rm to remove NA's

diff_solids <- water_stats$Solids - mu_solids # deviance between data frame and mean (X - xbar)

Sqr_solids <- diff_solids^2 # squared of the deviance (X - xbar)^2

SOS_solids <- sum(Sqr_solids) # Sum of squares using square of the deviance

var_solids <- var(water_stats$Solids) # calculates the variance

myvar_solids <- SOS_solids / length(water_stats$Solids) # calculates the variance from difference between SOS and number of scores(data points)

sd_solids <- sd(water_stats$Solids) # calculates sample the standard deviation
 
mu_solids <- sqrt(myvar_solids) # calculates the population standard deviation given by sigma

# The range calculated two ways:

range_solids <- diff(range(water_stats$Solids)) # calculates the range with the diff function
  
min_solids <- min(water_stats$Solids) # minimum value

max_solids <- max(water_stats$Solids) # maximum value

# range_turb <- max_turb - min_turb # range of data frame using min and max functions

```

## Plot the data to view linearity of the sample size.

```{r}
# Plot data to view linearity and whether data is normally distributed or not of sample set. 
plot(density(sample_water$Conductivity)) # Slightly bimodal peaks, indicative of underlying patterns. 
qqnorm(sample_water$Conductivity)

plot(density(sample_water$Solids)) # right skewness, normally distributed.
qqnorm(sample_water$Solids)

# Calculate Correlation matrix:
water_cor_sample <- round(cor(sample_water$Conductivity, sample_water$Solids, "complete.obs"), 4)
water_cor_sample # weak positive correlation between Conductivity and Solids

# Coefficient of Determination
water_cor_sample^2

# t critical value
tcrit <- qt(p = 0.975, df = 50-2)
tcrit

# Lower bound
#lb_sample <- 1.501 - tcrit*1.896
#lb_sample

# Upper bound
#ub_sample <- 1.501 + tcrit*1.896
#ub_sample


```

## Calculate the correlation and coefficient of determination value, then find the confidence interval boundaries of the sample size.

```{r}
# Calculate Correlation matrix:
water_cor_sample <- round(cor(sample_water$Conductivity, sample_water$Solids, "complete.obs"), 4)
water_cor_sample # weak positive correlation between Conductivity and Solids

# Coefficient of Determination
water_cor_sample^2

# t critical value
tcrit <- qt(p = 0.975, df = 50-2)
tcrit

# Lower bound
#lb_sample <- 1.501 - tcrit*1.896
#lb_sample

# Upper bound
#ub_sample <- 1.501 + tcrit*1.896
#ub_sample
```

## Create a linear regression model of the sample size with a least squares regression line and indications of slope and y intercept.

```{r}
# Create Linear Regression Model
Sample_Model <- lm(Solids ~ Conductivity, data = water_stats)
#Sample_Model$residuals
Sample_Model

# Store the coefficients on the linear model
my_intercept_sample <- round(Sample_Model$coefficients[1], 1)
my_slope_sample <- round(Sample_Model$coefficients[2], 1)

# Plot correlation.
plot(sample_water$Conductivity,
     sample_water$Solids,
     cex = 0.5, # make points smaller
     pch = 16, # make solid dots
     main = paste("LSRL: ", "Predicted y = ", my_intercept_sample, "+", my_slope_sample, "(x)"),
     xlab = "Conductivity of Water Bodies", 
     ylab = "Total Dissolved Solids")

# Plot LSRL
abline(Sample_Model, 
       lwd = 2,        # make the line wider
       col = "orange"     # color the line
       )

summary(Sample_Model)
```

## Is there a statistically significant correlation to access of safe drinking water between Conductivity and TDS (ppm) across local, regional, and national levels?

-   The $r^2$ value indicates a weak goodness of fit between the explanatory and response variables.

-   This could be due to ommited variables causing variance, high variability, or a high level of measurement error within the data.

-   The correlation itself indicates a weak positive correlation, and shows no linearity.

-   The LSRL indicates almost a completely horizontal slope. This shows a uniform rate and lack of change over time.

```{r}
# Plot data to view linearity and whether data is normally distributed or not. 
plot(density(water_stats$Conductivity)) # Slightly bimodal peaks, indicative of underlying patterns. 
qqnorm(water_stats$Conductivity)

plot(density(water_stats$Solids)) # right skewness, normally distributed.
qqnorm(water_stats$Solids)

# Calculate Correlation matrix:
water_cor <- round(cor(water_stats$Conductivity, water_stats$Solids, "complete.obs"), 4)
water_cor # weak positive correlation between Conductivity and Solids

# Coefficient of Determination
water_cor^2

# t critical value
tcrit <- qt(p = 0.975, df = 3276-2)
tcrit

# Lower bound
lb <- 1.501 - tcrit*1.896
lb

# Upper bound
ub <- 1.501 + tcrit*1.896
ub

# Create Linear Regression Model
Model <- lm(Solids ~ Conductivity, data = water_stats)
Model$residuals
Model

# Store the coefficients on the linear model
my_intercept <- round(Model$coefficients[1], 1)
my_slope <- round(Model$coefficients[2], 1)

# Plot correlation.
plot(water_stats$Conductivity,
     water_stats$Solids,
     cex = 0.5, # make points smaller
     pch = 16, # make solid dots
     main = paste("LSRL: ", "Predicted y = ", my_intercept, "+", my_slope, "(x)"),
     xlab = "Conductivity of Water Bodies", 
     ylab = "Total Dissolved Solids")

# Plot LSRL
abline(Model, 
       lwd = 2,        # make the line wider
       col = "red"     # color the line
       )

summary(Model)
```

-   Correlation is clustered points, positive value, weak. The population correlation is 0.0138, and the sample correlation is 0.0454. Both indicate positive, weak, and non linear.

-   Confidence interval boundaries (-2.216466, 5.218466)

-   p-value is very close to zero, hence smaller then $\alpha$ = 0.05, so we can reject the null hypothesis and state that there is statistical significance between the predictor and response variable, and that this correlation isn't by random chance alone.

-   We are 95% confident that the interval (-2.216, 5.218) captures the population slope of Total Dissolved Solids for Conductivity within local, regional, and national bodies of clean drinking water.

-   The population slope was 1.501, so yes, our interval captured the parameter.

## To what extent does total dissolved solids (TDS in ppm) concentrations statistically predict or correlate with conductivity levels across a random sample of water bodies (50 out of 3276)? (Does one predict the other?)

```{r}
# Set seed to 22.
#set.seed(22)

# Pick 4 different random samples
sampleIDs1 <- sample(1:3276, size = 50, replace = FALSE)
sampleIDs2 <- sample(1:3276, size = 50, replace = FALSE)
sampleIDs3 <- sample(1:3276, size = 50, replace = FALSE)
sampleIDs4 <- sample(1:3276, size = 50, replace = FALSE)

# Select the columns I'm interested in
myColumns <- c("Conductivity", "Solids")

#("pH", "Hardness", "Solids", "Chloramines", "Sulfate", "Conductivity", "Organic_carbon", "Trihalomethanes", "Turbidity", "Potability")

# Store the randomly selected rows on the given columns
sampleWater1 <- water_stats[sampleIDs1, myColumns]
sampleWater2 <- water_stats[sampleIDs2, myColumns]
sampleWater3 <- water_stats[sampleIDs3, myColumns]
sampleWater4 <- water_stats[sampleIDs4, myColumns]

# Calculate linear model for each sample
mod1 <- lm(water_stats$Solids~water_stats$Conductivity)
mod2 <- lm(water_stats$Solids~water_stats$Conductivity)
mod3 <- lm(water_stats$Solids~water_stats$Conductivity)
mod4 <- lm(water_stats$Solids~water_stats$Conductivity)

# Store the slope from each model
mod1slope <- round(mod1$coefficients[2], 1)
mod2slope <- round(mod2$coefficients[2], 1)
mod3slope <- round(mod3$coefficients[2], 1)
mod4slope <- round(mod4$coefficients[2], 1)

# Store r-squared for each model
mod1corr <- round(sqrt(summary(mod1)$r.squared), 2)
mod2corr <- round(sqrt(summary(mod2)$r.squared), 2)
mod3corr <- round(sqrt(summary(mod3)$r.squared), 2)
mod4corr <- round(sqrt(summary(mod4)$r.squared), 2)

# 2x2 viewing window
par(mfrow = c(2, 2))

# Plot 4 scatterplots with LSRL to illustrate sampling variability

plot(sampleWater1$Conductivity, sampleWater1$Solids, cex = 1.5, las = 1, pch = 16,  xlab = "Conductivity",  ylab = "TDS", main = paste("b =", mod1slope, "and r =", mod1corr))

abline(mod1, col = "red", lwd = 1.5)

plot(sampleWater2$Conductivity, sampleWater2$Solids, cex = 1.5, las = 1, pch = 16,  xlab = "Conductivity",  ylab = "TDS", main = paste("b =", mod2slope, "and r =", mod2corr))

abline(mod2, col = "red", lwd = 1.5)

plot(sampleWater3$Conductivity, sampleWater3$Solids, cex = 1.5, las = 1, pch = 16,  xlab = "Conductivity",  ylab = "TDS", main = paste("b =", mod3slope, "and r =", mod3corr))

abline(mod3, col = "red", lwd = 1.5)

plot(sampleWater4$Conductivity, sampleWater4$Solids, cex = 1.5, las = 1, pch = 16,  xlab = "Conductivity",  ylab = "TDS", main = paste("b =", mod4slope, "and r =", mod4corr))

abline(mod4, col = "red", lwd = 1.5)

# Note to self: making a function would have been wise here!
```
